---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# experDesign

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/llrs/experDesign.svg?branch=master)](https://travis-ci.org/llrs/experDesign)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/llrs/experDesign?branch=master&svg=true)](https://ci.appveyor.com/project/llrs/experDesign)
[![Coverage status](https://codecov.io/gh/llrs/experDesign/branch/master/graph/badge.svg)](https://codecov.io/github/llrs/experDesign?branch=master)
[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Project Status: Active - The project has reached a stable, usable state and is being actively developed.](http://www.repostatus.org/badges/latest/active.svg)](http://www.repostatus.org/#active)
<!-- badges: end -->

The goal of experDesign is to help you decide which samples go in which batch, 
reducing the potential batch bias when analyzing.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

```r
# install.packages("devtools")
devtools::install_github("llrs/experDesign")
```
## Example

This is a basic example which shows you how to solve a common problem:

```{r design}
library("experDesign")
metadata <- expand.grid(height = seq(60, 80, 5), 
                        weight = seq(100, 300, 50),
                        sex = c("Male","Female"))
head(metadata, 15)
d <- design(metadata, 24)
# It is a list but we can convert it to a vector with:
batch_names(d)

```

Naively one would either fill some batches fully or distribute them not evenly 
(the  first 17 packages toghether, the next 17 and so on). This solution ensures 
that the data is randomized.

If you need space for replicates to control for batch effect you can use:

```{r replicates}
r <- replicates(metadata, 24, 5)
r
```

Which seeks as controls the most diverse values and adds them to the samples 
distribution.

# Previous work

The [OSAT](https://bioconductor.org/packages/OSAT/) package handles categorical 
variables but not numeric data. It doesn't work with our data.

The [minDiff](https://github.com/m-Py/minDiff) package reported in [Stats.SE](https://stats.stackexchange.com/a/326015/105234), handles both 
numeric and categorical data. But it can only optimize for two nominal criteria.
It doesn't work for our data.

The [DeclareDesign](https://github.com/DeclareDesign/DeclareDesign) or [OSAT](http://bioconductor.org/packages/OSAT) packages are also relevant

Question in [Bioinformatics.SE](https://bioinformatics.stackexchange.com/q/4765/48)

# Other

Please note that this project is released with a [Contributor Code of Conduct](CODE_OF_CONDUCT.md).
By participating in this project you agree to abide by its terms.
