---
title: "exprDesign"
author: "Lluís Revilla Sancho"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    fig_caption: true
    code_folding: show
    self_contained: yes
    toc_float:
      collapsed: true
      toc_depth: 3
vignette: >
  %\VignetteIndexEntry{exprDesign}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitsetup, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_knit$set(root.dir = ".")
knitr::opts_chunk$set(collapse = TRUE, warning = TRUE)
BiocStyle::markdown()
library("BiocStyle")
set.seed(445)
library("experDesign")
```

# Introduction

This package was developed to help prepare some samples to be send to a 
facility. It assumes that you have collected the samples and the information 
but you still need to do the experiment in several batches due to technical or 
practical limitations. 

# Previous work

Before building this package I would like to give credit to those that made 
also efforts in this direction:

The [OSAT](https://bioconductor.org/packages/OSAT/) package handles categorical 
variables but not numeric data.

The [minDiff](https://github.com/m-Py/minDiff) package reported in [Stats.SE](https://stats.stackexchange.com/a/326015/105234), handles both 
numeric and categorical data, but it can only optimize for two nominal criteria.

# Approach

Of all the space of possible combinations of samples, it looks for the
combination which minimizes the differences between each subgroup

 - If the variable is categorical it looks so that the entropy is closer to random.
 - If the variable is numeric it looks so that the distribution is closer to the original.
 - If there are `NA` (not available values) it looks to distribute them randomly.

# Preparation

We can use the survey dataset for the examples:
```{r}
data(survey, package = "MASS") 
head(survey)
```

The dataset has numeric, categorical values and some `NA`'s value.

# Workflow

Imagine that we can only work in groups of 70, and we want to randomize by Sex, 
Smoke, Age, and by writing hand.  
There are `r choose(237, 70)` combinations some of them would be have in a 
single experiment all the right handed, we can't measure all these combinations
but we can find an optimum value.

```{r, fig.show='hold'}
omit <- c("Wr.Hnd", "NW.Hnd", "Fold", "Pulse", "Clap", "Exer", "Height", "M.I")
(keep <- colnames(survey)[!colnames(survey) %in% omit])
index <- design(pheno = survey, size.batch = 70, omit = omit)
```
To convert to a value to append to the data.frame we can use:
```{r batch, eval=FALSE}
batch <- batch_names(index)
```

Or if we want to do so automatically we can do:
```{r}
df <- inspect(index, survey, omit = omit)
head(df)
```

If a data.frame has a column with name "batch" we can use the function 
distribution to see if it is randomly distributed. Or we can see the mean 
difference between each subgroup and the original value of the data.frame.

```{r}
evaluate_entropy(index, survey)
evaluate_independence(index, survey)

evaluate_na(index, survey)

evaluate_mean(index, survey)
evaluate_sd(index, survey)
evaluate_mad(index, survey)
```

The numbers provided are just an indicator of how far are they from a perfect 
distance. The farthest and index is, the worst it is to account for batch effect.

If we want the original numbers we can use `evaluate_index`, and look for specific values:
```{r}
ev <- evaluate_index(index, survey)
ev["entropy", "Sex",]
ev[1:4, "Age",]
evaluate_independence(index, survey)
```

We can see that the sex is fairly distributed and that the mean age ratio is 
almost perfectly distributed. But we can compare with the original distribution 
using `evaluate_orig`:

```{r}
orig <- evaluate_orig(survey)
orig[, "Age"]
```
We can see that the `sd` is slightly out, but the mean and mad for each subset 
is fairly similar to the original points.

# Unbalanced setting

In this case the data was fairly balanced (check it out in the `orig` object) 
but let's create an unbalanced dataset to check it.

```{r unbalanced}
n <- 99
samples <- 100
unbalanced <- data.frame(Classroom = rep(c("A", "B"), each = samples/2),
                         Sex = c(rep("M", n), rep("F", samples-n)))
table(unbalanced)
```
In this dataset the classroom a single classroom has all the females (`r 50 -n`).

```{r}
i <- design(unbalanced, 15)

# Mean entropy en each subset
rowMeans(evaluate_index(i, unbalanced)["entropy", , ])
# Original entropy on the dataset
evaluate_orig(unbalanced)["entropy", ]
# Dispersion of the entropy
apply(evaluate_index(i, unbalanced)["entropy", , ], 1, sd)
```
We can see that in this simple case where a single variable has all the the other cases we approximately reached the same entropy levels. We can avoid batch factor when the data

# Quality check

If you need a subset with the samples that are more diverse you can use the following functions:

```{r QC}
size <- optimum_size(survey)
samples <- extreme_cases(survey, size = size)
```


# SessionInfo

```{r}
sessionInfo()
```

