---
title: "exprDesign"
author: "Lluís Revilla Sancho"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    fig_caption: true
    code_folding: show
    self_contained: yes
    toc_float:
      collapsed: true
      toc_depth: 3
vignette: >
  %\VignetteIndexEntry{exprDesign}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r knitsetup, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_knit$set(root.dir = ".")
knitr::opts_chunk$set(collapse = TRUE, warning = TRUE)
BiocStyle::markdown()
library("BiocStyle")
set.seed(445)
library("experDesign")
```

# Introduction

This package was developed to help prepare some samples to be send to a 
facility. It assumes that you have collected the samples and the information 
but you still need to do the experiment in several batches due to technical or 
practical limitations. The question that tries to answer is: 

> Which samples are done in which batch?

Of all the possible combinations of samples, it looks for the combination which 
minimizes the differences between each subgroup according to the following rule:

 - If the variable is categorical it looks so that the entropy is closer to random.
 - If the variable is numeric it looks so that the distribution is closer to the original.
 - If there are `NA` (not available values) it looks to distribute them randomly.

# Previous work

Before building this package I would like to give credit to those that made 
also efforts in this direction:

The [OSAT](https://bioconductor.org/packages/OSAT/) package handles categorical 
variables but not numeric data.

The [minDiff](https://github.com/m-Py/minDiff) package reported in [Stats.SE](https://stats.stackexchange.com/a/326015/105234), handles both 
numeric and categorical data, but it can only optimize for two nominal criteria.

This package handles both numerical and categorical variables as many as you 
wish. But the accuracy depends on the number of `iterations` done. 
But if you only have data falling on any of these condition the results will be 
more accurate.

# Preparation

We can use the survey dataset for the examples:
```{r}
data(survey, package = "MASS") 
head(survey)
```

The dataset has numeric, categorical values and some `NA`'s value.

# Picking samples for each batch

Imagine that we can only work in groups of 70, and we want to randomize by Sex, 
Smoke, Age, and by writing hand.  
There are `r choose(237, 70)` combinations some of them would be have in a 
single experiment all the right handed students. We could measure all these combinations
but we can try to find an optimum value.

```{r, fig.show='hold'}
# To reduce the variables used:
omit <- c("Wr.Hnd", "NW.Hnd", "Fold", "Pulse", "Clap", "Exer", "Height", "M.I")
(keep <- colnames(survey)[!colnames(survey) %in% omit])
head(survey[, keep])

# Looking for groups at most of 70 samples.
index <- design(pheno = survey, size_subset = 70, omit = omit)
index
```

We can transform then into a vector to append to the file or to pass to the lab mate with:
```{r batch_names}
batch_names(index)
```

# Checking results

We can check the values of the sampling with:

```{r evaluate_index}
out <- evaluate_index(index, survey[, keep])
out[, "Age", ]
```

And we can compare with the original distribution using `evaluate_orig`:

```{r evaluate_orig}
orig <- evaluate_orig(survey)
orig[, "Age"]
```


# Unbalanced setting

In the previous case the data was mostly balanced (check it out in the `orig` object) 
but let's create an unbalanced dataset to check it.

```{r unbalanced}
n <- 99
samples <- 100
unbalanced <- data.frame(Classroom = rep(c("A", "B"), each = samples/2),
                         Sex = c(rep("M", n), rep("F", samples-n)),
                         Age = rnorm(samples, mean = 25, sd = 3))
head(table(unbalanced))
```

In this dataset the classroom a single classroom has all the females (`r 50 -n`).

```{r}
i <- design(unbalanced, 15)

# Mean entropy en each subset
rowMeans(evaluate_index(i, unbalanced)["entropy", , ])
# Original entropy on the dataset
evaluate_orig(unbalanced)["entropy", ]
# Dispersion of the entropy
apply(evaluate_index(i, unbalanced)["entropy", , ], 1, sd)
```
We can see that in this simple case where a single variable has all the other cases we approximately reached the same entropy levels. 

# Quality check

If you need a subset with the samples that are more diverse you can use the 
following function:

```{r QC}
samples <- extreme_cases(survey, size = 10)
survey[samples, ]
```

It is recommended to add replicates on the several batches to estimate the differences between the batches. If you want to add replicates on the batches you can use the following:

```{r replicates}
# Looking for groups at most of 70 samples.
index_replicates <- replicates(pheno = survey, size_subset = 70, 
                               controls = 10, omit = omit)
index_replicates
```

The controls will be those on all the batches:

```{r}
survey[Reduce(intersect, index_replicates), ]
```

If you have biological replicates use the information you have about them, 
but if you have technical replicates do not add them to the table used by 
`design` or `replicates`.

# Utilities

To convert to a value to append to the data.frame we can use:
```{r batch}
batch <- batch_names(index)
```

Or if we want to do so automatically we can do:
```{r}
df <- inspect(index, survey, omit = omit)
head(df)
```

# Comparing groups

If a data.frame has a column with name "batch" we can use the function 
distribution to see if it is randomly distributed. Or we can see the mean 
difference between each subgroup and the original value of the data.frame.

```{r}
evaluate_entropy(index, survey)
evaluate_independence(index, survey)

evaluate_na(index, survey)

evaluate_mean(index, survey)
evaluate_sd(index, survey)
evaluate_mad(index, survey)
```

The numbers provided are just an indicator of how far are they from a perfect 
distance. The farthest and index is, the worst it is to account for batch effect.

If we want the original numbers we can use `evaluate_index`, and look for specific values:
```{r}
ev <- evaluate_index(index, survey)
ev["entropy", "Sex",]
ev[1:4, "Age",]
evaluate_independence(index, survey)
```

# SessionInfo

```{r}
sessionInfo()
```

